Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex
More polishing in intro

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens,let},
emphstyle=\upshape\sffamily\color{dkred}, 
columns=fullflexible,
escapechar=\#,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape, }

@@

{\em Quotient lenses}~\cite{quotientlenses} loosen the strong restrictions of
bijective string lenses by including equivalence
relations on the source and target formats and requiring that the
round-tripping lens laws 
hold  modulo these equivalences, rather than ``on the nose.'' Quotient lenses 
make it possible to synchronize many more data formats than is possible with
plain bijective lenses\bcp{Do we justify this claim later?}.

The bidirectional programming language {\em
Boomerang}~\cite{boomerang} is the only language we know of that
implements a general class of quotient lenses.  Unfortunately, these
lenses to have two main drawbacks.

The first drawback is that defining an equivalence
relation in Boomerang requires specifying a {\em canonizer}, which is
a pair of functions $f$ and $g$ where $f$
normalizes data from the
external format by choosing canonical representations
and $g$ provides an
inverse by specifying the external form of each canonical
representation. 
The disadvantage of this approach is that writing
a quotient lens in Boomerang requires defining four
different data formats: two ``external'' formats which specify the data to
be transformed by the lens, and two ``internal'' formats which specify the
canonical representations for the source and target data.

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.  As an example, consider the following QRE for writing
author names:
\iflastminute
\bcp{I'd love to use a visible space symbol for important space characters
  -- this would make the code examples very much clearer!  But it will
  require a bit of hacking: We can use showstringspaces=false}
\fi
%
\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

The second drawback with writing quotient lenses in Boomerang is that
even after specifying the four required data formats, the programmer must
still define a lens for converting between the canonical
representations.   Experience shows that writing such lenses can be
tedious (consider having to rearrange the order of data items by recursively
using operators to \kw{swap} adjacent fields)  and that satisfying the type
checker can be difficult because many seemingly useful lenses contain
ambiguities disallowed by the Boomerang type checker~\cite{optician}.

@@

\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
\end{lstlisting}
binds \codefont{l} to a synthesized QRE lens mapping between names in
the comma-separated form describd by the QRE \codefont{comma\_name}
and the space-separated form described by the
QRE \codefont{space\_name}. 
The lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Our synthesized quotient lenses have a relatively
restricted form (which greatly speeds up synthesis), but the 
normalization theorem means 
this restriction does not compromise expressiveness.

The following list summarizes the contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs), which are
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses}, which are lenses that translate
  between data formats specified using QREs. (Section~\ref{QRE-lenses}).
  \item We descibe a particular set of QRE lens combinators and prove
  that QRE lenses defined using these combinators can be normalized
  (Theorem~\ref{normal form}), which is the main technical
  contribution of this paper.
  (Section~\ref{QRE-lenses}).
  \item We use this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses
  previously studied in~\cite{optician}(Section~\ref{synth}).
  \item We extend Boomerang with QREs and QRE lens synthesis. 
  We use the resulting implementation to demonstrate the utility of
  our approach by synthesizing QRE lenses between
  data formats drawn from the Optician benchmark suite (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

let bib_to_end : bibtex $\Leftrightarrow$ end_note = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\iflastminute
\bcp{a possible readability suggestion: define a named constant for a quote
  character.  (Not sure if this makes sense, or if it improves things...)}
\fi}
\label{fig:example-lens}
\end{figure}

>>>>

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens,let},
emphstyle=\upshape\sffamily\color{dkred}, 
columns=fullflexible,
escapechar=\#,
mathescape, 
xleftmargin=1.5em,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
}

@@

{\em Quotient lenses}~\cite{quotientlenses} loosen the strong restrictions of
bijective string lenses by including equivalence
relations on the source and target formats and requiring that the
round-tripping lens laws 
hold  modulo these equivalences, rather than ``on the nose.'' Quotient lenses 
make it possible to synchronize many more data formats than is possible with
plain bijective lenses\bcp{Do we justify this claim later?}. 
%
The bidirectional programming language {\em
Boomerang}~\cite{boomerang} 
implements a general class of quotient lenses.  Unfortunately, these
lenses to have two main drawbacks.  \bcp{This wording may raise a question
  in the reader's mind about whether this paper is really about quotient
  lenses in general or whether it is specifically a proposal to tweak
  Boomerang.  I.e., ``if I am not a Boomerang user, do I want to bother
  reading this paper?''  My answer would be yes, because quotient
  lenses and QREs are interesting in themselves.  But I don't think we've
  made this point very convincingly in the way we've framed the paper.}

The first shortcoming is that defining an equivalence
relation in Boomerang requires explicitly providing a {\em canonizer}, 
a pair of functions $(f,g)$ where $f$
normalizes data from the
external format by choosing canonical representations
and $g$ provides an
inverse by specifying the external form of each canonical
representation. 
The disadvantage of this approach is that writing
a quotient lens in Boomerang requires separately defining four
different data formats: two ``external'' formats which specify the data to
be transformed by the lens, and two ``internal'' formats which specify the
canonical representations for the source and target data\bcp{Plus writing
  two canonizers!}.

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.  As an example, consider the following QRE for writing
author names:
%
\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

The second drawback with writing quotient lenses in Boomerang is that even
after specifying the four required data formats, the programmer must still
define a lens for converting between the canonical representations.
Experience shows that writing such lenses can be tedious (e.g., consider
having to rearrange the order of data items by recursively using operators
to \kw{swap} adjacent fields) and that satisfying the type checker can be
difficult because the most obvious way to express a lens runs afoul of one
or more ``ambiguity checks'' that the type checker performs in order to make
sure that the lens laws hold~\cite{optician}.

@@

\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
\end{lstlisting}
binds \codefont{l} to a synthesized QRE lens mapping between names in the
comma-separated form describd by the QRE \codefont{comma\_name} and the
space-separated form described by the QRE \codefont{space\_name}.  The
lenses generated by our synthesis algorithm are guaranteed to translate data
back and forth between source and target QREs and to satisfy the provided
examples.  Our synthesized quotient lenses have a relatively restricted
form, where the sequential composition operator is not used and all
canonization occurs at the very edges (which greatly speeds up synthesis),
but the normalization theorem means this restriction does not compromise
expressiveness.

In summary, our main contributions are:
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs), 
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses}, which translate between data formats
  specified using QREs (Section~\ref{QRE-lenses}).
  \item We design a specific set of useful QRE lens combinators and prove
  that QRE lenses defined using these combinators can be put into a
  restricted normal form (Section~\ref{QRE-lenses}).  This is
  the main technical contribution of this paper.
  \item We leverage this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses, which was
  previously studied in~\cite{optician} (Section~\ref{synth}).
  \item We extend Boomerang with QREs and QRE lens synthesis and demonstrate
  the utility and practicality of our approach by synthesizing QRE lenses
  between a variety of data formats drawn from the Optician benchmark suite
  (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} present related and future work.

@@

let bib_to_end : bibtex $\Leftrightarrow$ end_note = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{A plain bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions (which are omitted for brevity).  
\iflastminute
\bcp{a possible readability suggestion: define a named constant for a quote
  character.  (Not sure if this makes sense, or if it improves things...)}
\fi}
\label{fig:example-lens}
\end{figure}
